Liveness-Based Garbage Collection for Lazy Languages

### 2015-06-02 paper/paper.tex
added algo back
fix for compile error

\documentclass[9pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage [plain,noend,noline,boxed]{algorithm2e}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
%\usepackage{program}
\usepackage{algorithmic}

\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it 
      %probably doesn't scale anyway.
\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
\newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\blankout}[1]{}

@@

Functional  programming  languages  depend  on  garbage  collector  to
efficiently  reclaim  unused   memory  allocated  by  programs  either
explicitly  (for  example using  constructors)  or implicitly  (during
creation    of    closures).     However,   empirical    studies    on
Haskell~\cite{rojemo96lag},  Scheme~\cite{karkare06effectiveness}  and
Java~\cite{shaham02estimating}   programs  have  shown   that  garbage
collectors leave uncollected a lot of memory objects that are not live
(here {\em live}ness  means the object can potentially  be used by the
program at  a later  stage).  Due  to lack of  a suitable  analysis to
distinguish  reachable  objects  from  live objects,  current  garbage
collectors conservatively approximate the  liveness of heap objects by
their reachability from a set of memory locations called the {\em root
  set\/}. This results in unnecessary memory retention.

Programs written in languages having lazy semantics for evaluation put
an further demands on  memory as they require heap allocated
closures to be  carried from the point of  occurrence of an expression
to the point of (first) real use of that expression.
%% This   may   be   mitigated  by   strictness
%% analysis~\cite{some-refs} that  can statically determine  the patterns
%% of future usage  of an expression, thus allowing  some of the closures
%% to be  evaluated at the point  of the occurrence  of the corresponding
%% expression.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% MOTIVATING EXAMPLE
\newcommand{\nilfigure}
{\scalebox{0.75}{
\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
\begin{pspicture}(0,0)(1,1)
%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(1,1)
\putnode{start}{origin}{0}{0}{}
\putnode{stop}{origin}{10}{10}{}
\ncline[offsetB=0,nodesepB=0,linewidth=.7]{-}{start}{stop} %here
\end{pspicture}
}}

@@

\caption{Example Program and its Memory Graph.}\label{fig:mot-example}    
%  \end{center}
%%%%}\color{Myblue} end
\end{figure*}
We propose the  use of liveness analysis of heap  cells and strictness
analysis  of expressions  to  reduce the  memory  requirement of  lazy
functional  programming languages.   We modify  garbage collectors  to
preserve  only {\em  live} data---a  subset of  {\em reachable  data}.
This results in an increase  in the garbage reclaimed and consequently
in fewer  garbage collections.  
Using  a central notion  of demand,  we formulate  a context-sensitive
liveness  analysis  and  prove   it  correct.   We  use  a  0-CFA-like
conservative   approximation   to   annotate   each   allocation   and
function-call program point  with a finite-state automaton---which the
garbage-collector inspects to curtail reachability during marking.

While this idea has been shown to be effective~\cite{asati14lgc} for a
first-order  eager  language,  a  straightforward  extension  of  this
earlier  technique  is  not   possible  in  lazy  languages,  where  a
heap-allocated   object  may  either   be  a   evaluated  data   or  a
closure. Since liveness is brought about by evaluation of closures and
given the  uncertainty in lazy  languages regarding the  program point
where this  evaluation takes  place, this introduces  complications in
the analysis itself.  Further,  since we want to extend liveness-based
collection to closures, we  track closures  to their
points of creation, from where we obtain their liveness. Since this is
done  at  runtime,  this  requires considerable  modification  to  the
earlier GC scheme.

%% We  previously  proposed an  interprocedural  liveness analysis  to
%% improve   garbage   collection   first   order  eager   subset   of
%% Scheme~\cite{asati14lgc}.   The current paper  extends the  work to
%% handle lazy evaluation semantics for the same language.

Experiments  confirm  the expected  performance  benefits.  The  early
evaluation   of   closures  decreases   the   memory   demands  of   a
program. Further,  liveness-based collection results in  a increase in
garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
collections,  a  (some quantification)  decrease  in  the memory  size
required to run programs,  and reduced overall garbage collection time
for a majority of programs.

@@

\subsubsection{Organization of the paper}
Section~\ref{sec:defs} gives
the syntax and semantics of
the language used to illustrate our analysis along with basic concepts
and    notations.
% Section~\ref{sec:operational} describes the operational semantics of
% the language.
Liveness analysis is  described in Section~\ref{sec:liveness} followed
by  a  sketch  of  a  correctness proof  relative  to  a  non-standard
semantics.  Section~\ref{sec:computing}  shows how to  encode liveness
as   finite-state  automata.    Section~\ref{sec:experiments}  reports
experimental  results  and Section~\ref{sec:lgc-always-better}  proves
that a liveness based collector  can never do more garbage collections
than a reachability based collector.
% and Section~\ref{sec:conclusion} concludes the paper.

@@

The body  of a function ${\mathit  f}$ will be  denoted as $e_{\mathit
  f}$. for  analysis purposes it is convenient  to regard $e_\mainpgm$
as part of a function definition $(\DEFINE\ ({\tt main})\ e_\mainpgm)$
as in  C\@.  We  write $\pi\!:\!e$ to  associate the label  $\pi$ (not
part  of the  language  syntax)  with the  program  point just  before
expression $e$.

@@

A variable is {\em live} if  there is a possibility of its value being
used in  future computations  and dead if  it is definitely  not used.
Heap-allocated data needs a richer model of liveness which talks about
liveness  of  pointers  including  fields  of  $\CONS$  cells.   Using
$\acar$, $\acdr$  to represent access using $\CAR$  and $\CDR$ fields,
the  liveness  of the  structure  reachable  from  a variable  can  be
represented by a set of {\em access paths} i.e.  prefix-closed strings
from $\{\acar,\acdr\}^\ast$.   As an  example, if $x$  is a  list with
liveness      $\{\epsilon,     \acdr,      \acdr\acar,     \acdr\acdr,
\acdr\acdr\acar\}$, then future computations  can only refer up to the
second and  third members of $x$.   A {\em liveness  environment} is a
mapping from variables to subsets of $\{\acar,\acdr\}^\ast$, but often
expressed  as a set,  for example  by writing  $\{x.\epsilon, x.\acdr,
x.\acdr\acdr,        y.\epsilon\}$        instead        of        $[x
  \mapsto\{\epsilon,\acdr,\acdr\acdr\},           y\mapsto\{\epsilon\},
  z\mapsto\{\}]$.    In  this   notation,  $x   \mapsto  \{\epsilon\}$
represents access using $x$ itself and $x \mapsto \{\}$ indicates that
$x$ is  dead.  We associate liveness environments  with program points
(denoted  by  $\pi$ and  $\psi$).   The  liveness  at $\pi:e$  is  the
liveness just before executing $e$.

A  notion  related  to  liveness  is  {\em  demand}.   While  liveness
represents the  future use of variables, demands  represent the future
use of the  values of expressions. The demand on  an expression $e$ is
again a  set of  access paths---that subset  of $\{\acar,\acdr\}^\ast$
which the context of $e$ may explore of $e$'s result.  To see the need
for  demands,  consider   the  let  expression  $\pi:(\LET~x\leftarrow
(\CDR~y)~\IN~  \pi':\SRETURN~x)$.  Assume  that  the  context  of  this
expression produces  the liveness $[x \mapsto  \lbrace \epsilon, \acar
  \rbrace]$ at $\pi'$.  Due to the \LET\ definition which binds $x$ to
$(\CDR~y)$, the  liveness of $x$ at  $\pi'$ now becomes  the demand on
$(\CDR~y)$.    This,  in   turn,  generates   the   liveness  $\lbrace
y.\epsilon,  y.\acdr,  y.\acdr\acar  \rbrace$  at $\pi$  (the  set  of
$y$-rooted  accesses  required  to  explore $\lbrace  \epsilon,  \acar
\rbrace$  paths of  the result  of  $(\CDR~y)$)\footnote{The classical
  analogue  is strong  liveness analysis,  where  $y$ and  $z$ have  a
  liveness of  $\lbrace\epsilon\rbrace$ at  the entry $n:  x:=y+z$, if
  and only if  $x$ has a liveness of  $\lbrace\epsilon\rbrace$ at exit
  of $n$.   In our terminology, the  liveness of $x$ at  exit from $n$
  becomes the  demand on $y+z$.  This, in turn,  generates the  liveness $\lbrace
  y.\epsilon, z.\epsilon \rbrace$ at the entry of $n$.}.
%% Exploring $\lbrace \epsilon, \acar \rbrace$ of the
%% result of $(\CDR~y)$ first requires access using $y$ ( and the \CDR\ field of the cons cell pointed by
%% $y$.

Note that  liveness refers to  access using variables and  fields, and
not  to $\CONS$  cells (i.e.\  to edges  in the  memory graph,  not to
locations themselves). 
\begin{figure*}[t!]
  \begin{tabular}{cc}
    \begin{boxedminipage}{.45\textwidth}
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{.27cm} $\pi_3\!\!:\,
            $(\LET\ \pv\ $\leftarrow 0$ \IN
            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\, (\SRETURN~\psi_2:\pv)$
            \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{.34cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{.34cm} $\pi_7\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\, (\SRETURN~\psi_3:\pz)$)))))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $(\DEFINE\ (\main)$
          \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow$ \ldots  \IN
	  \UNL{3}   \hspace*{.05cm}$\pi_{10}\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{4}   \hspace*{.15cm}    $\pi_{11}\!\!:\,
          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
          \UNL{5}  \hspace*{.25cm}  $\pi_{12}\!\!:\, (\SRETURN~\psi_4:\pw)))))$
\end{uprogram}}
        %%} 
    \end{boxedminipage}

    &

    \begin{boxedminipage}{.51\textwidth}

        \small
        \begin{tabular}{lc}
        $\mathcal{L}(e_{\length},\sigma,\Lfonly) = $ 
            \begin{tabular}{lc}            
              $[~ep_1 \mapsto \lbrace \px.\sigma_{\length}, \xl.\cbar\sigma_{\length} \rbrace ,~ep_2 \mapsto \lbrace \rbrace , $\\
              $~~ep_3 \mapsto \lbrace \pz.\sigma_{\length},
                \py.\cbar\sigma_{\length}, \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, $\\
              $~~~~~~~~~~~~~~\;\;
                \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}
                \rbrace~\rbrack$ \\
            \end{tabular} \\
            $\cal{P_{\mbox{\tiny \length}}} = $
              \begin{tabular}{lc}
                $~[
                  ~\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
                $ \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
                  \rbrace, $\\
$ ~~~\pi_3 \mapsto \lbrace ep_2
                  \rbrace $, $ \pi_4 \mapsto \lbrace ep_2
                  \rbrace$, $\pi_5 \mapsto \lbrace  ep_3 \rbrace$, \\
                $~~~\pi_6 \mapsto \lbrace ep_3 \rbrace,  \pi_7 \mapsto
                  \lbrace ep_3\rbrace, \pi_8 \mapsto \lbrace ep_3 \rbrace~\rbrack$
              \end{tabular}\\
              $\mathcal{L}(e_{\main},\sigma,\Lfonly) = $ 
              \begin{tabular}{lc}            
                $[$
                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}$, $\pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}$, $\ldots  \rbrace $ \
                  $]$
              \end{tabular} \\
              $\cal{P_{\mbox{\tiny \main}}} = $
                \begin{tabular}{lc}
                  $~\lbrack$~$\pi_6 \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_7 \mapsto \lbrace ep_4 \rbrace, $ \
                  $\pi_8 \mapsto \lbrace ep_4 \rbrace,~\psi_4 \mapsto \lbrace ep_4 \rbrace~\rbrack$  
                \end{tabular}
        \end{tabular}

    \end{boxedminipage} 
    %% \end{ tabular}}
  \end{tabular}
%\end{picture}
\kern -3ex
\vspace{0.5in}
\caption{Example Program and its liveness maps.{\color{red} Fix the
    formatting of the table on the right}}\label{fig:mot-example2}    
\end{figure*}

@@

@@

 In each  rule defining $\mathit{ref}$, the  null demand ($\emptyset$)
 generates no  liveness.  Notice that this is  different from liveness
 of  eager  languages,  where,   for  instance,  (\CDR~$x$)  would  be
 evaluated  even for  the null  demand. This  would require  an access
 using $x$ resulting in  $x.\epsilon$ being live.  A non-null demand of
 $\sigma$ on (\CDR~$x$), is transformed to the liveness $\{x.\epsilon,
 x.\acdr\sigma\}$.  In an opposite  sense, the demand of $\acdr\sigma$
 on  (\CONS~$y$~$z$) is  transformed to  the demand  $\sigma$  on $z$.
 Note that \CONS\ does not,  by itself, dereference its arguments. The
 rules for (\PRIM~x~y) and  (\NULLQ~x) are on similar lines. Constants
 generate no liveness.

For  the  case  of a  user-function  call  we  use a  third  parameter
$\Lfonly$  that  represents the  summaries  of  all  functions in  the
program.   $\Lfonly_{\mathit f}$  (the  component of  $\Lfonly$ for  a
specific  function $f$),  expresses  how the  demand  $\sigma$ on  the
result  of a  call to  $f$  is transformed  into the  liveness of  its
parameters at the  beginning of the call.  $\Lfonly$  is determined by
the judgement  form $\mathit{Prog} \len \Lfonly$  using inference rule
({\sc live-define}).  This rule  describes the fixed-point property to
be satisfied  by $\Lfonly$, namely, the  demand transformation assumed
for each  function in  the program  should be the  same as  the demand
transformation  calculated from  their  bodies.  As  we  shall see  in
Section~\ref{sec:grammar-formulation},  we  convert  the rule  into  a
grammar and  the least  solution satisfying the  rule is  the language
generated by this  grammar.  The least solution also  ensures the safe
collection of the greatest amount of garbage.

We next  describe the function $\mathcal{L}$  that propagates liveness
across expressions.  Note that,  unlike eager languages, the syntactic
occurrence  of an expression  in a  program denotes  the point  of 
creation of its closure, not necessarily its evaluation.
To see the  consequences of this, consider an analysis  of the body of
$\length$ with a demand $\sigma$.  Clearly, this is also the demand on
the return value $z$ giving the liveness at the program point $\pi_8$
as $\{\pz.\sigma\}$.  The liveness at $\pi_8$ is propagated backwards
as in traditional liveness analysis.

Since   \pz\  is   $(1  +   \py)$,  the   liveness  of   $\sigma$  for
\pz\ translates into a demand  of $\sigma$ on $(1+\py)$, which in turn
generates a  liveness of $\{\py.\epsilon\}$ at  $\pi_7$.  We now make
the following observations:
\begin{enumerate}
\item While the liveness of \py\  should be killed at $\pi_6$ since it
  is  beyond the scope  of the  definition of  \py, we  do not  do so.
  However, this  does not cause any  imprecision in the way  of the GC
  marking more cells as live,  since, during execution, \py\ is yet to
  come  into  existence at  $\pi_6$,  and  its  liveness will  not  be
  consulted by the garbage collector at this point.
\item In an eager language, \py\ would not have been live at $\pi_8$,
  a program point  beyond its last use.  In  a lazy language, however,
  the  definition $\pz  \leftarrow  (1+\py)$, does  not  result in  an
  evaluation  of  $(1+\py)$;  instead   a  closure  is  created.   The
  evaluation  actually takes  place while  evaluating $(\SRETURN~\pz)$.
  We therefore  record  the liveness of $\py$ at the program  
  point $\pi_8$,  indeed at every  program point from $\pi_8$  up to
  the beginning  of the program $\pi_1$ (since  we do not  kill liveness).
\end{enumerate}

We thus regard  the body of the function as a  tree and consider paths
from  the  root   of  this  tree  to  program   points  which  trigger
evaluation\footnote{Program   points  which  trigger   evaluation  are
  labelled  with   $\psi$  {\color  {red}  ---slightly   late  in  the
    day}}---\SIF\ conditions  and \SRETURN\ expressions.   We call such
paths {\em e-paths}.   For the function \length, there  are three such
paths  $ep_1   =  \langle   \pi_1,  \pi_2,  \psi_1\rangle$,   $ep_2  =
\langle\pi_1, \pi_2, \pi_3, \pi_4,  \psi_2\rangle$ and $ep_3 = \langle
\pi_1, \pi_2, \pi_5, \pi_6, \pi_7, \pi_8, \psi_3\rangle$.

>>>>

\documentclass[10pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage {algorithmic}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
%\usepackage{program}
\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it 
      %probably doesn't scale anyway.
\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
\newcommand{\comment}[1]{{\color{Myblue}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\blankout}[1]{}

@@

Functional  programming  languages  depend  on  garbage  collector  to
efficiently  reclaim  unused  memory   allocated  by  programs  either
explicitly (for example using constructors) or implicitly (for example
for      closures).       However,      empirical      studies      on
Haskell~\cite{rojemo96lag},  Scheme~\cite{karkare06effectiveness}  and
Java~\cite{shaham02estimating}  programs   have  shown   that  garbage
collectors leave uncollected a lot of memory objects that are not live
(here {\em live}ness  means the object can potentially be  used by the
program at  a later  stage).  Due  to lack of  a suitable  analysis to
distinguish  reachable  objects  from live  objects.  current  garbage
collectors conservatively approximate the  liveness of heap objects by
their reachability from a set of memory locations called the {\em root
  set\/} resulting in unnecessary memory usage.

Languages having lazy  semantics for evaluation put  further demand on
memory as they  require heap allocated closure to be  carried from the
point of occurrence  of an expression to the point  of (first) real use
of  that expression.  Strictness analysis~\cite{some-refs}  statically
determines  the  patterns  of  future usage  of  an  expression,  thus
allowing some  of the  closures to  be evaluated at  the point  of the
occurrence of the corresponding expression.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% MOTIVATING EXAMPLE
\newcommand{\nilfigure}
{\scalebox{0.75}{
\psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
\begin{pspicture}(0,0)(1,1)
%\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(1,1)
\putnode{start}{origin}{0}{0}{}
\putnode{stop}{origin}{10}{10}{}
\ncline[offsetB=0,nodesepB=0,linewidth=.7]{-}{start}{stop} %here
\end{pspicture}
}}

@@

\caption{Example Program and its Memory Graph.}\label{fig:mot-example}    
%  \end{center}
%%%%}\color{Myblue} end
\end{figure*}
We propose the  use of liveness analysis of heap  cells and strictness
analysis  of expressions  to  reduce the  memory  requirement of  lazy
functional  programming languages.   We modify  garbage collectors  to
preserve  only {\em  live} data---a  subset of  {\em reachable  data}.
This results in an increase  in the garbage reclaimed and consequently
in fewer  garbage collections.  Further, we eagerly evaluate closures
that are guaranteed to be evaluated in the future (strict closures).

The main contribution  of this paper is an uniform  framework for {\em
  liveness} and {\em strictness}  analysis to enable the optimizations
mentioned  above. Using  a central  notion of  demand, we  formulate a
context-sensitive liveness  analysis for structured data  and prove it
correct.  We then  present strictness as a dual of  liveness. We use a
0-CFA-like conservative approximation to  annotate each allocation and
function-call program point with  a finite-state automaton---which the
garbage-collector inspects to curtail reachability during marking. The
result of  the strictness analysis,  being necessarily finite,  can be
directly used to insert code to evaluate closures.

We previously proposed an interprocedural liveness analysis to improve
garbage collection first order eager subset of Scheme~\cite{cc-paper}.
The current paper extends the work to handle lazy evaluation semantics
for the same language.

Experiments  confirm  the  expected  performance benefits.  The  early
evaluation   of   closures  decreases   the   memory   demands  of   a
program. Further,  liveness-based collection results in  a increase in
garbage  reclaimed  and  a   consequent  decrease  in  the  number  of
collections, a decrease  in the memory size required  to run programs,
and  reduced  overall  garbage  collection  time  for  a  majority  of
programs.

@@

\subsubsection{Organization of the paper}
Section~\ref{sec:defs} gives
the syntax and semantics of
the language used to illustrate our analysis along with basic concepts
and    notations.
% Section~\ref{sec:operational}   describes    the
% operational semantics  of the  language.
Liveness analysis is described in
Section~\ref{sec:liveness} followed by a sketch of a correctness
proof relative to a non-standard semantics.
Section~\ref{sec:computing} shows how to encode liveness as
finite-state automata.
Section~\ref{sec:experiments}   reports   experimental   results and
Section~\ref{sec:lgc-always-better} proves that a liveness based
collector can never do more garbage collections than a reachability
based collector. 
% and Section~\ref{sec:conclusion} concludes the paper.

@@

The body of the program is the expression denoted by $e_\mainpgm$; for
analysis purposes it is convenient to regard $e_\mainpgm$ as part of a
function definition $(\DEFINE\ ({\tt main})\ e_\mainpgm)$ as in C\@.
We write  $\pi\!:\!e$ to  associate the label  $\pi$ (not part  of the
language syntax) with the program point just before expression $e$.

@@

A variable is {\em live} if  there is a possibility of its value being
used  in future computation  and dead  if it  is definitely  not used.
Heap-allocated data needs a richer model of liveness which talks about
liveness  of  pointers  including  fields  of  $\CONS$  cells.   Using
$\acar$, $\acdr$  to represent access using $\CAR$  and $\CDR$ fields,
the  liveness  of the  structure  reachable  from  a variable  can  be
represented by a set of {\em access paths} i.e.  prefix-closed strings
from $\{\acar,\acdr\}^\ast$.   As an  example, if $x$  is a  list with
liveness      $\{\epsilon,     \acdr,      \acdr\acar,     \acdr\acdr,
\acdr\acdr\acar\}$, then future computations  can only refer up to the
second and  third members of $x$.   A {\em liveness  environment} is a
mapping from variables to subsets of $\{\acar,\acdr\}^\ast$, but often
expressed as  a set, for example by  writing $\{x.\epsilon, x.\acdr, x.\acdr\acdr, 
y.\epsilon\}$               instead               of               $[x
  \mapsto\{\epsilon,\acdr,\acdr\acdr\},     y\mapsto\{\epsilon\},
  z\mapsto\{\}]$.   In  this   notation,   $x  \mapsto   \{\epsilon\}$
represents access using $x$ itself and $x \mapsto \{\}$ indicates that
$x$ is dead.  We associate  a liveness environment with program points
$\pi$.  The liveness at $\pi:e$  is the liveness just before executing
$e$.

A  notion related  to  liveness is  {\em  demand}.  The  demand on  an
expression  $e$  is again  a  set  of  access paths---that  subset  of
$\{\acar,\acdr\}^\ast$ which  the context of $e$ may  explore of $e$'s
result.   To see  the need  for demands,  consider the  let expression
$\pi:(\LET~x\leftarrow (\CDR~y)~\IN~ \pi':\SRETURN~x)$. Assume that the
context of  this expression produces the liveness  $[x \mapsto \lbrace
  \epsilon, \acar  \rbrace]$ at $\pi'$.   Due to the  \LET\ definition
which  binds $x$  to $(\CDR~y)$,  the liveness  of $x$  at  $\pi'$ now
becomes  the  demand on  $(\CDR~y)$.   This,  in  turn, generates  the
liveness $\lbrace y.\epsilon,  y.\acdr, y.\acdr\acar \rbrace$ at $\pi$
(the set of $y$-rooted accesses required to explore $\lbrace \epsilon,
\acar  \rbrace$  paths  of  the  result  of  $(\CDR~y)$)\footnote{The
  classical analogue  is strong liveness  analysis, where $y$  and $z$
  have  a  liveness  of  $\lbrace\epsilon\rbrace$  at  the  entry  $n:
  x:=y+z$,    if   and    only   if    $x$   has    a    liveness   of
  $\lbrace\epsilon\rbrace$ at  exit of  $n$.  In our  terminology, the
  liveness of $x$  at exit from $n$ becomes the  demand on $y+z$. This
  generates the  liveness $\lbrace y.\epsilon,  z.\epsilon \rbrace$ at
  the entry of $n$.}.
%% Exploring $\lbrace \epsilon, \acar \rbrace$ of the
%% result of $(\CDR~y)$ first requires access using $y$ ( and the \CDR\ field of the cons cell pointed by
%% $y$.
Note that  liveness refers to  access using variables and  fields, and
not  to $\CONS$  cells (i.e.\  to edges  in the  memory graph,  not to
locations themselves).

@@

%% Finally, we use $\Lfonly$  to range over {\em demand transformers}.
%% These transform  demands on  a function call  into demands  for its
%% formal  parameters:   given  a   user  function  $f$,   defined  by
%% $(\DEFINE\ (f\  x_1\ \ldots\ x_n)\  \ e_f)$ and called  with demand
%% $\sigma$,  then $\Lf{f}{i}{\sigma}$  is  the liveness  of $x_i$  at
%% $e_f$.

@@

 In each  rule defining $\mathit{ref}$, the  null demand ($\emptyset$)
 generates no  liveness.  Notice that this is  different from liveness
 of  eager  languages,  where,   for  instance,  (\CDR~$x$)  would  be
 evaluated even for the null  demand, and this would require an access
 using  $x$.   A  non-null   demand  of  $\sigma$  on  (\CDR~$x$),  is
 transformed  to  the  liveness  $\{x.\epsilon, x.\acdr\sigma\}$.   In  an
 opposite  sense, the  demand of  $\acdr\sigma$ on  (\CONS~$y$~$z$) is
 transformed to  the demand  $\sigma$ on $z$.   Note that  \CONS\ does
 not, by itself, dereference  its arguments. The rules for (\PRIM~x~y)
 and (\NULLQ~x) are on similar lines. Constants generate no liveness.

For  the  case  of a  user-function  call  we  use a  third  parameter
$\Lfonly$  that  represents the  summaries  of  all  functions in  the
program.  $\Lfonly_f$  (the  component  of $\Lfonly$  for  a  specific
function $f$),  expresses how the demand  $\sigma$ on the  result of a
call to $f$ is transformed into  the liveness of its parameters at the
beginning of the call.  $\Lfonly$  is determined by the judgement form
$\mathit{Prog} \len \Lfonly$ using inference rule ({\sc live-define}).
This  rule  describes the  fixed-point  property  to  be satisfied  by
$\Lfonly$, namely, the demand transformation assumed for each function
in  the  program should  be  the  same  as the  demand  transformation
calculated    from   their    bodies.     As   we    shall   see    in
Section~\ref{sec:grammar-formulation},  the least  solution satisfying
the  rule  is obtained  by  converting  the  rule into  an  equivalent
grammar.  The least solution also  ensures that the greatest amount of
garbage is collected.

%%   Analogously to classical liveness being computed as a solution of
%% dataflow   equations,  we   require,  via   inference   rule  ({\sc
%% live-define}), $\Lfonly$ to  satisfy the fixed-point property that:
%% when we  assume $\Lfonly$ to  be the family of  demand transformers
%% for the program then the  calculated liveness of each function body
%% $\mathcal{L}(e_f,\sigma,\Lfonly)$    agrees   with    the   assumed
%% $\Lfonly$.   As  usual,  there  are  often  multiple  solutions  to
%% $\Lfonly$; all are  safe (see Section~\ref{sec:correctness}) but we
%% prefer  the least  one  as  giving the  least  liveness subject  to
%% safety---and hence greatest amount of garbage collected.

We next  describe the function $\mathcal{L}$  that propagates liveness
across  expressions.   This brings  us  to  the  second way  in  which
liveness analysis for lazy languages differ from eager languages---the
point  where  an  expression   gets  evaluated  cannot  be  statically
determined.  To  deal with this, we introduce  flow insensitivity into
our  analysis.
